---
title: "Statistical Analysis of Porfolios"
author: "Econ 424"
date: "July 2023"
output: slidy_presentation
---

## Instructions
Please read Zivot Ch.15.  Below are sample codes to carry out the analyses mentioned in the chapter.

Your task is to:

1. Motivate (why we care) and illustrate (how it's done) the impact of statistical or estimation uncertainties in model parameters on the portfolio analysis results.

2. Carry out the analyses on the dataset you used for HW4/HW5 by applying the above sample codes below.

Please note that the bootstrap procedure will be explained by Group 2 (who will go first, so you only need to briefly describe what's being done and focus on the results instead.)

3. Discuss your findings.  

- Specifically, address how estimation errors in the GWN model parameters affect the efficient portfolios we obtain. e.g. how are the means, standard deviations, correlations, and furthermore, weights on the global minimum variance portfolio, Sharpe ratios, tangency portfolios affected?

- As such, what are some implications?  (e.g. should one be more confident about efficient portfolios with targeted means or S.D? other thoughts?  how about changing sample periods, how robust are your conclusions?)


## Set options and load packages

```{r warning=FALSE,message=FALSE}
options(digits=3, width=70)
library(ellipse)
library(IntroCompFinR)
library(mvtnorm)
library(PerformanceAnalytics)
library(zoo)
library(corrplot)
Sys.setenv(TZ="UTC")
cex.val = 2
```

## Two Asset Example

Use two asset example data from Introduction to Portfolio Theory chapter re-scaled for monthly time horizon:

```{r}
mu.A = 0.175/12
sig.A = 0.258/sqrt(12)
sig2.A = sig.A^2
mu.B = 0.055/12
sig.B = 0.115/sqrt(12)
sig2.B = sig.B^2
rho.AB = -0.164
sig.AB = rho.AB*sig.A*sig.B
sigma.mat = matrix(c(sig2.A, sig.AB, sig.AB, sig2.B), 2, 2, byrow=TRUE) 
mu.vec = c(mu.A, mu.B) 
sd.vec = c(sig.A, sig.B) 
names(mu.vec) = names(sd.vec) = c("asset.A", "asset.B") 
dimnames(sigma.mat) = list(names(mu.vec), names(mu.vec))
```

```{r}
mu.vec
sd.vec
sig.AB
```

Consider an equally weighted portfolio of assets A and B. The GWN model parameters for this portfolio are:

```{r}
x1.A = 0.5 
x1.B = 1 - x1.A 
x1.vec = c(x1.A, x1.B) 
names(x1.vec) = names(mu.vec)
mu.p1 = x1.A*mu.A + x1.B*mu.B 
sig2.p1 = x1.A^2 * sig2.A + x1.B^2 * sig2.B + 2*x1.A*x1.B*sig.AB 
sig.p1 = sqrt(sig2.p1)
cbind(mu.p1, sig2.p1, sig.p1)
```

## Risk-return trade offs for example data

```{r}
plot(sd.vec, mu.vec, pch=16, col="black", 
     ylim=c(0, 0.02), xlim=c(0, 0.1),
     xlab=expression(sigma[p]), ylab=expression(mu[p]),
     cex=2, cex.lab = 1.5)
abline(h=0)
points(sig.p1, mu.p1, pch=16, col="black", cex=2)
text(x=sig.A, y=mu.A, labels="A", pos=4, cex = 1.5)
text(x=sig.B, y=mu.B, labels="B", pos=4, cex = 1.5)
text(x=sig.p1, y=mu.p1, labels="P", pos=4, cex = 1.5)
```


## Simulate hypothetical returns from GWN model

We use the above GWN model parameters to simulate $T = 60$ hypothetical returns for assets A and B and the equally weighted portfolio:

```{r}
library(mvtnorm)
n.obs = 60 
set.seed(125) 
returns.sim = rmvnorm(n.obs, mean=mu.vec, sigma=sigma.mat) 
colnames(returns.sim) = names(mu.vec)
rp1.sim = returns.sim%*%x1.vec
```

## GWN model estimates on simulated data

The estimates of the GWN model parameters from the simulated returns for the individual assets are:

```{r}
muhat.vals = colMeans(returns.sim) 
sig2hat.vals = apply(returns.sim, 2, var) 
sigmahat.vals = apply(returns.sim, 2, sd) 
covmat.hat = cov(returns.sim) 
covhat = covmat.hat[1,2]
corhat = covhat/(prod(sigmahat.vals))
rbind(muhat.vals, sig2hat.vals, sigmahat.vals)
c(covhat, corhat)
```

The estimated standard errors and 95% confidence intervals are:

```{r}
# estimated standard errors
se.muhat = sigmahat.vals/sqrt(n.obs) 
se.sigmahat = sigmahat.vals/sqrt(2*n.obs) 
rbind(se.muhat, se.sigmahat)
# 95% confidence intervals 
lower.mu = muhat.vals - 2*se.muhat 
upper.mu = muhat.vals + 2*se.muhat 
lower.sigma = sigmahat.vals - 2*se.sigmahat 
upper.sigma = sigmahat.vals + 2*se.sigmahat 
cbind(lower.mu, upper.mu) 
cbind(lower.sigma, upper.sigma)
```

* The means are not estimated as precisely as the volatilities
* All of the true values for $\mu$ and $\sigma$ are contained in the 95% confidence intervals.

## GWN model estimates for equally weighted portfolio

For the equally weighted portfolio, the GWN model estimates computed directly from the simulated portfolio returns are:

```{r}
muhat.p1 = mean(rp1.sim) 
sig2hat.p1 = as.numeric(var(rp1.sim)) 
sigmahat.p1 = sd(rp1.sim) 
cbind(muhat.p1, sig2hat.p1, sigmahat.p1)
```

The portfolio estimates computed from the asset estimates give the same results:

```{r}
muhat.p1 = as.numeric(x1.A*muhat.vals[1] + x1.B*muhat.vals[2]) 
sig2hat.p1 = as.numeric(x1.A^2 * sig2hat.vals[1] + x1.B^2 * sig2hat.vals[2] + 2*x1.A*x1.B*covhat) 
sighat.p1 = sqrt(sig2hat.p1) 
cbind(muhat.p1, sig2hat.p1, sigmahat.p1)
```

The estimated standard errors and 95% confidence intervals are:

```{r}
# estimated standard errors
se.muhat.p1 = sigmahat.p1/sqrt(n.obs) 
se.sigmahat.p1 = sigmahat.p1/sqrt(2*n.obs) 
cbind(se.muhat.p1, se.sigmahat.p1)
# 95% confidence intervals 
lower.mu = muhat.p1 - 2*se.muhat.p1 
upper.mu = muhat.p1 + 2*se.muhat.p1 
lower.sigma = sigmahat.p1 - 2*se.sigmahat.p1 
upper.sigma = sigmahat.p1 + 2*se.sigmahat.p1 
ans = rbind(c(lower.mu, upper.mu),             
c(lower.sigma, upper.sigma)) 
colnames(ans) = c("lower", "upper") 
rownames(ans) = c("mu.p", "sigma.p") 
ans
```

* As with the individual assets, the portfolio mean is estimated less precisely than the portfolio volatility and the true values are in the 95% confidence intervals.

## True and estimated risk-return tradeoffs

```{r}
plot(sd.vec, mu.vec, pch=16, col="black",       
     ylim=c(-0.01, 0.04), xlim=c(0, 0.1),      
     xlab=expression(sigma[p]), ylab=expression(mu[p]),      
     cex=2) 
abline(h=0) 
points(sig.p1, mu.p1, pch=16, col="black", cex=2) 
text(x=sig.A, y=mu.A, labels="A", pos=4, cex = 1.5) 
text(x=sig.B, y=mu.B, labels="B", pos=4, cex = 1.5) 
text(x=sig.p1, y=mu.p1, labels="P", pos=4, cex = 1.5) 
# estimates 
points(sigmahat.vals, muhat.vals, pch=16, col="blue", cex=2) 
points(sigmahat.p1, muhat.p1, pch=16, col="blue", cex=2) 
text(x=sigmahat.vals[1], y=muhat.vals[1], labels="A Hat",       
     col="blue", pos=4, cex = 1.5) 
text(x=sigmahat.vals[2], y=muhat.vals[2], labels="B Hat",       
     col="blue", pos=4, cex = 1.5) 
text(x=sigmahat.p1, y=muhat.p1, labels="P Hat",       
     col="blue", pos=4, cex = 1.5)
```

* B and B Hat are close, B Hat is slightly below B
* P and P Hat are moderately close, P Hat is above P
* A and A Hat are not close, A Hat is high above A

## True and estimated risk-return tradeoffs with 95% confidence ellipses

```{r}
library(ellipse)
plot(sd.vec, mu.vec, pch=16, col="black",       
     ylim=c(-0.01, 0.06), xlim=c(0, 0.1),      
     xlab=expression(sigma[p]), ylab=expression(mu[p]),cex=2) 
abline(h=0) 
points(sig.p1, mu.p1, pch=16, col="black", cex=2) 
text(x=sig.A, y=mu.A, labels="A", pos=4, cex = 1.5) 
text(x=sig.B, y=mu.B, labels="B", pos=4, cex = 1.5) 
text(x=sig.p1, y=mu.p1, labels="P", pos=4, cex = 1.5) 
# estimates points(sigmahat.vals, muhat.vals, pch=16, col="blue", cex=2) 
points(sigmahat.p1, muhat.p1, pch=16, col="blue", cex=2) 
text(x=sigmahat.vals[1], y=muhat.vals[1], labels="A Hat",       
   col="blue", pos=4, cex = 1.5) 
text(x=sigmahat.vals[2], y=muhat.vals[2], labels="B Hat",       
     col="blue", pos=4, cex = 1.5) 
text(x=sigmahat.p1, y=muhat.p1, labels="P Hat",       
     col="blue", pos=4, cex = 1.5)
# Create asymptotic variances 
V.A = matrix(c(se.sigmahat[1]^2, 0 ,0, se.muhat[1]^2), 2, 2, byrow=TRUE) 
V.B = matrix(c(se.sigmahat[2]^2, 0 ,0, se.muhat[2]^2), 2, 2, byrow=TRUE) 
V.P = matrix(c(se.sigmahat.p1^2, 0 ,0, se.muhat.p1^2), 2, 2, byrow=TRUE)
# plot confidence ellipses 
lines(ellipse(V.A, centre=c(sigmahat.vals[1], muhat.vals[1]), level=0.95),         
      col="red" , lwd=2) 
lines(ellipse(V.B, centre=c(sigmahat.vals[2], muhat.vals[2]), level=0.95),         
      col="red", lwd=2 ) 
lines(ellipse(V.P, centre=c(sigmahat.p1, muhat.p1), level=0.95),         
      col="red", lwd=2 )
```

* The `ellipse()` function from the R package **ellipse** is used to draw the 95% confidence sets
* The confidence ellipses are longer in the $\mu$ direction
* The confidence ellipse for $(\sigma_{A}, \mu_{A})$ is much bigger than the other ellipses

## True and estimated risk-return tradeoffs with bootstrap estimates

The following R code creates $B = 500$ bootstrap estimates of $(\sigma_{i},\mu_{i})$ for $i = A,B,P$:

```{r}
n.boot = 500 
mu.boot = matrix(0, n.boot, 3) 
sd.boot = matrix(0, n.boot, 3) 
colnames(mu.boot) = colnames(sd.boot) = c("A", "B", "P")
set.seed(123) 
for (i in 1:n.boot) {   
   boot.idx = sample(n.obs, replace=TRUE)   
   ret.boot = returns.sim[boot.idx, ]    
   rp1.boot = ret.boot%*%x1.vec   
   ret.boot = cbind(ret.boot, rp1.boot)   
   mu.boot[i, ] = colMeans(ret.boot)   
   sd.boot[i, ] = apply(ret.boot, 2, sd)  
}
```

## True and estimated risk-return tradeoffs with bootstrap estimates 

```{r}
# show estimates with 95% confidence intervals 
plot(sd.vec, mu.vec, type="n",  ylim=c(-0.01, 0.06), xlim=c(0, 0.1),      
     xlab=expression(sigma[p]), ylab=expression(mu[p])) 
# bootstrap estimates 
points(sd.boot[, "A"], mu.boot[, "A"], pch=16, col="lightblue") 
points(sd.boot[, "B"], mu.boot[, "B"], pch=16, col="grey") 
points(sd.boot[, "P"], mu.boot[, "P"], pch=16, col="pink") 
# true values points(sd.vec, mu.vec, pch=16, col="black", cex=2) 
abline(h=0) 
points(sig.p1, mu.p1, pch=16, col="black", cex=2) 
text(x=sig.A, y=mu.A, labels="A", pos=4, cex = 1.5) 
text(x=sig.B, y=mu.B, labels="B", pos=4, cex = 1.5) 
text(x=sig.p1, y=mu.p1, labels="P", pos=4, cex = 1.5) 
# estimates points(sigmahat.vals, muhat.vals, pch=16, col="blue", cex=2) 
points(sigmahat.p1, muhat.p1, pch=16, col="blue", cex=2) 
text(x=sigmahat.vals[1], y=muhat.vals[1], labels="A Hat",       
    col="blue", pos=4, cex = 1.5) 
text(x=sigmahat.vals[2], y=muhat.vals[2], labels="B Hat",       
    col="blue", pos=4, cex = 1.5) 
text(x=sigmahat.p1, y=muhat.p1, labels="P Hat",       
     col="blue", pos=4, cex = 1.5) 
# plot confidence ellipses 
lines( ellipse(V.A, centre=c(sigmahat.vals[1], muhat.vals[1]), level=0.95),         
      col="red" , lwd=2) 
lines( ellipse(V.B, centre=c(sigmahat.vals[2], muhat.vals[2]), level=0.95),         
      col="red", lwd=2 ) 
lines( ellipse(V.P, centre=c(sigmahat.p1, muhat.p1), level=0.95),         
     col="red", lwd=2 ) 
```

## True and estimated portfolio frontier

Here, we create the true and estimated portfolio frontiers for the following portfolios:

```{r}
x.A = seq(from=-0.4, to=1.4, by=0.1) 
x.B = 1 - x.A
```

The risk-return characteristics of the portfolios on the true portfolio frontier are:

```{r}
mu.p = x.A*mu.A + x.B*mu.B 
sig2.p = x.A^2 * sig2.A + x.B^2 * sig2.B + 2*x.A*x.B*sig.AB 
sig.p = sqrt(sig2.p)
```

The risk-return characteristics of the portfolios on the estimated frontier are:

```{r}
muhat.p = x.A*muhat.vals["asset.A"] + x.B*muhat.vals["asset.B"] 
sig2hat.p = x.A^2 * sig2hat.vals["asset.A"] + x.B^2 * sig2hat.vals["asset.B"] + 2*x.A*x.B*covhat 
sighat.p = sqrt(sig2hat.p)
```

## True and estimated portfolio frontier

```{r}
plot(sig.p, mu.p, pch=16, type="b", col="black",       
     ylim=c(-0.01, 0.041), xlim=c(0, 0.125),      
     xlab=expression(sigma[p]), ylab=expression(mu[p]), cex=1) 
abline(h=0) 
points(sd.vec, mu.vec, pch=16, col="black", cex=2) 
text(x=sig.A, y=mu.A, labels="A", pos=3, cex = 1.5) 
text(x=sig.B, y=mu.B, labels="B", pos=3, cex = 1.5) 
# estimated assets and frontier 
points(sighat.p, muhat.p, type="b", pch=16, col="blue", cex=1) 
points(sigmahat.vals, muhat.vals, pch=16, col="blue", cex=2) 
text(x=sigmahat.vals[1], y=muhat.vals[1], labels="A Hat",       
     col="blue", pos=3, cex = 1.5) 
text(x=sigmahat.vals[2], y=muhat.vals[2], labels="B Hat",       
     col="blue", pos=2, cex = 1.5)
```

* Due to the large estimation error in $\hat{\mu}_{A}$, the estimated frontier is considerably higher than the true frontier. 
* As a result, portfolios on the estimated frontier appear to have higher reward-to-risk properties than they actually do.

## True and estimated portfolio frontier with bootstrap estimates

* Sampling uncertainty about the estimated frontier can be easily computed using the bootstrap. 
* The process is the same as in the case of a single asset. 
* For each bootstrap sample, estimate the expected return and volatility of each portfolio on the frontier and then plot these bootstrap pairs on the plot showing the estimated frontier from the sample returns.

```{r}
# initialize matrices 
n.boot = 1000 
mu.boot = matrix(0, n.boot, length(x.A)) 
sd.boot = matrix(0, n.boot, length(x.A)) 
colnames(mu.boot) = colnames(sd.boot) = paste("P", 1:length(x.A), sep=".") 
# bootstrap loop 
set.seed(123) 
for (i in 1:n.boot) {   
   boot.idx = sample(n.obs, replace=TRUE)   
   ret.boot = returns.sim[boot.idx, ]    
# GWN model estimates   
   muhat.boot = colMeans(ret.boot)   
   sig2hat.boot = apply(ret.boot, 2, var)   
   sigmahat.boot = sqrt(sig2hat.boot)   
   covhat.boot = cov(ret.boot)[1,2]   
# portfolio risk return estimates   
   mu.boot[i, ] = x.A*muhat.boot[1] + x.B*muhat.boot[2]   
   sig2.boot = x.A^2 * sig2hat.boot[1] + x.B^2 * sig2hat.boot[2] + 2*x.A*x.B*covhat.boot   
   sd.boot[i, ] = sqrt(sig2.boot) 
}
```

## True and estimated portfolio frontier with bootstrap estimates

```{r}
# set up plot area
plot(sig.p, mu.p, type="n", ylim=c(-0.02, 0.07), xlim=c(0, 0.13),      
     xlab=expression(sigma[p]), ylab=expression(mu[p])) 
# plot bootstrap estimates 
for (i in 1:length(x.A)) {   
   points(sd.boot[, i], mu.boot[, i], pch=16, col="lightblue") 
}
# plot true frontier
points(sig.p, mu.p, pch=16, type="b", col="black", cex=1) 
abline(h=0) 
# plot true assets
points(sd.vec, mu.vec, pch=16, col="black", cex=2) 
text(x=sig.A, y=mu.A, labels="A", pos=3, cex = 1.5) 
text(x=sig.B, y=mu.B, labels="B", pos=3, cex = 1.5) 
# plot estimated frontier 
points(sighat.p, muhat.p, type="b", pch=16, col="blue", cex=1) 
# plot estimated assets
points(sigmahat.vals, muhat.vals, pch=16, col="blue", cex=2) 
text(x=sigmahat.vals[1], y=muhat.vals[1], labels="A Hat",       
     col="blue", pos=3, cex = 1.5) 
text(x=sigmahat.vals[2], y=muhat.vals[2], labels="B Hat",       
     col="blue", pos=2, cex = 1.5)
```

* The true frontier portfolios are the black dots, the estimated frontier portfolios are the dark blue dots and the bootstrap risk-return pairs are the light blue dots. 
* The bootstrap estimates form a light blue cloud around the estimated frontier. 
* The bootstrap cloud can be interpreted as approximating a confidence interval around estimated frontier.  

## True and estimated global minimum variance portfolio 

The true and estimated global minimum variance portfolio are:

```{r}
# True global min variance portfolio weights
xA.min = (sig2.B - sig.AB)/(sig2.A + sig2.B - 2*sig.AB) 
xB.min = 1 - xA.min 
c(xA.min, xB.min)
# Estimated min variance portfolio weights
xA.min.hat = (sig2hat.vals[2] - covhat)/(sig2hat.vals[1] + sig2hat.vals[2] - 2*covhat) 
xB.min.hat = 1 - xA.min.hat 
c(xA.min.hat, xB.min.hat)
```

The true and estimated expected return and volatility of the global minimum variance portfolio are:

```{r}
# Expected return and volatility of true global minimum variance portfolio 
mu.p.min = xA.min*mu.A + xB.min*mu.B 
sig2.p.min = xA.min^2 * sig2.A + xB.min^2 * sig2.B + 2*xA.min*xB.min*sig.AB 
sig.p.min = sqrt(sig2.p.min) 
c(mu.p.min, sig.p.min)
# Expected return and volatility of estimated global minimum variance portfolio 
mu.p.min.hat = as.numeric(xA.min.hat*muhat.vals[1] + xB.min.hat*muhat.vals[2]) 
sig2.p.min.hat = xA.min.hat^2 * sig2hat.vals[1] + xB.min.hat^2 * sig2hat.vals[2] + 2*xA.min.hat*xB.min.hat*covhat 
sig.p.min.hat = as.numeric(sqrt(sig2.p.min.hat)) 
c(mu.p.min.hat, sig.p.min.hat)
```

## True and estimated global minimum variance portfolio 

```{r, echo=FALSE}
plot(sig.p, mu.p, pch=16, type="b", col="black",       
ylim=c(-0.01, 0.041), xlim=c(0, 0.125),      
xlab=expression(sigma[p]), ylab=expression(mu[p]), cex=1) 
abline(h=0) 
points(sd.vec, mu.vec, pch=16, col="black", cex=2) 
text(x=sig.A, y=mu.A, labels="A", pos=3, cex = 1.5) 
text(x=sig.B, y=mu.B, labels="B", pos=3, cex = 1.5) 
# estimated assets and frontier 
points(sighat.p, muhat.p, type="b", pch=16, col="blue", cex=1) 
points(sigmahat.vals, muhat.vals, pch=16, col="blue", cex=2) 
text(x=sigmahat.vals[1], y=muhat.vals[1], labels="A Hat",       
col="blue", pos=3, cex = 1.5) 
text(x=sigmahat.vals[2], y=muhat.vals[2], labels="B Hat",       
col="blue", pos=1, cex = 1.5) 
# true min variance portfolio 
points(sig.p.min, mu.p.min, pch=16, col="green", cex=2) 
text(x=sig.p.min, y=mu.p.min, labels="GMIN", col="green", pos=2, cex=1.5) 
# estimated min variance portfolio 
points(sig.p.min.hat, mu.p.min.hat, pch=16, col="orange", cex=2) 
text(x=sig.p.min.hat, y=mu.p.min.hat, labels="GMIN Hat", col="orange", pos=2, cex=1.5) 
```

## Bootstrapping the global minimum variance portfolio estimates

To compute $B = 1000$ bootstrap estimates of the global minimum variance portfolio parameters use:

```{r}
# initialize matrices 
n.boot = 1000 
weights.boot = matrix(0, n.boot, 2) 
stats.boot = matrix(0, n.boot, 2) 
colnames(weights.boot) = names(mu.vec)  
colnames(stats.boot) = c("mu", "sigma") 
# bootstrap loop 
set.seed(123) 
for (i in 1:n.boot) {   
   boot.idx = sample(n.obs, replace=TRUE)   
   ret.boot = returns.sim[boot.idx, ]    
# GWN model estimates   
   muhat.boot = colMeans(ret.boot)   
   sig2hat.boot = apply(ret.boot, 2, var)   
   sigmahat.boot = sqrt(sig2hat.boot)   
   covhat.boot = cov(ret.boot)[1,2]   
# global minimum variance portfolio weights   
   weights.boot[i, 1] = (sig2hat.boot[2] - covhat.boot)/(sig2hat.boot[1] + sig2hat.boot[2] - 2*covhat.boot)   
   weights.boot[i, 2] = 1 - weights.boot[i, 1]   
# portfolio risk return estimates   
   stats.boot[i, "mu"] = weights.boot[i, 1]*muhat.boot[1] + weights.boot[i, 2]*muhat.boot[2]   
   sig2.boot = (weights.boot[i, 1]^2 * sig2hat.boot[1] + weights.boot[i, 2]^2 * sig2hat.boot[2]                 
                + 2*weights.boot[i, 1]*weights.boot[i, 2]*covhat.boot)   
   stats.boot[i, "sigma"] = sqrt(sig2.boot) 
}
```

## Statistical properties of estimated global minimum variance portfolio weights

The bootstrap bias estimates for the global minimum variance portfolio weights are:

```{r}
colMeans(weights.boot) - c(xA.min.hat, xB.min.hat)
```

* These values are close to zero suggesting that the estimated weights are unbiased. 

The bootstrap standard error estimates for the weights are:

```{r}
apply(weights.boot, 2, sd)
```

* The standard errors estimates are identical and close to 0.05 which is not too large.

## bootstrap distribution of global minimum variance weights

```{r}
par(mfrow=c(2,2))   
   hist(weights.boot[, 1], main="Asset A", xlab="Asset A weights", col="slateblue1")   
   abline(v=xA.min.hat, col="white", lwd=2) 
   hist(weights.boot[, 2], main="Asset B", xlab="Asset B weights", col="slateblue1")   
   abline(v=xB.min.hat, col="white", lwd=2)
   qqnorm(weights.boot[, 1])   
   qqline(weights.boot[, 1])   
   qqnorm(weights.boot[, 2])   
   qqline(weights.boot[, 2]) 
par(mfrow=c(1,1))
```

* These distributions are centered at the sample estimates (white vertical lines) and look slightly asymmetric

## Statistical properties of estimated global minimum variance portfolio mean and volatility

The bootstrap estimates of bias for the estimated expected return and volatility of the global minimum variance portfolio are:

```{r}
colMeans(stats.boot) - c(mu.p.min.hat, sig.p.min.hat)
```

* These small values indicate that the estimates are roughly unbiased. 

The bootstrap standard error estimates are:

```{r}
apply(stats.boot, 2, sd)
```

* Here, the bootstrap standard error estimate for $\hat{\mu}_{p,\mathrm{min}}$ is larger than the estimate for $\hat{\sigma}_{p,\mathrm{min}}$ indicating more uncertainty about the mean of the global minimum variance portfolio than the volatility of the portfolio.
  
## bootstrap distribution of global minimum variance weights

```{r}
par(mfrow=c(2,2))   
   hist(stats.boot[, 1], main="mu", xlab="bootstrap estimates of mu", col="slateblue1")   
   abline(v=mu.p.min.hat, col="white", lwd=2)   
   hist(stats.boot[, 2], main="sigma", xlab="bootstrap estimates of sigma", col="slateblue1")   
   abline(v=sig.p.min.hat, col="white", lwd=2)   
   qqnorm(stats.boot[, "mu"])   
   qqline(stats.boot[, "mu"])   
   qqnorm(stats.boot[, "sigma"])   
   qqline(stats.boot[, "sigma"]) 
par(mfrow=c(1,1))
```

## Bootstrapping the global minimum variance portfolio estimates

```{r, echo=FALSE}
plot(sig.p, mu.p, pch=16, type="n", ylim=c(-0.01, 0.041), xlim=c(0, 0.125),      
     xlab=expression(sigma[p]), ylab=expression(mu[p])) 
# bootstrap estimates 
points(stats.boot[, "sigma"], stats.boot[, "mu"], pch=16, col="lightgoldenrod2") 
# true frontier 
points(sig.p, mu.p, pch=16, type="b", col="black", cex=1) 
abline(h=0) 
points(sd.vec, mu.vec, pch=16, col="black", cex=2) 
text(x=sig.A, y=mu.A, labels="A", pos=3, cex = 1.5) 
text(x=sig.B, y=mu.B, labels="B", pos=3, cex = 1.5) 
# estimated assets and frontier 
points(sighat.p, muhat.p, type="b", pch=16, col="blue", cex=1) 
points(sigmahat.vals, muhat.vals, pch=16, col="blue", cex=2) 
text(x=sigmahat.vals[1], y=muhat.vals[1], labels="A Hat",       
     col="blue", pos=3, cex = 1.5) 
text(x=sigmahat.vals[2], y=muhat.vals[2], labels="B Hat",       
     col="blue", pos=1, cex = 1.5) 
# true min variance portfolio 
points(sig.p.min, mu.p.min, pch=16, col="green", cex=2) 
text(x=sig.p.min, y=mu.p.min, labels="GMIN", col="green", pos=2, cex=1.5) 
# estimated min variance portfolio 
points(sig.p.min.hat, mu.p.min.hat, pch=16, col="orange", cex=2) 
text(x=sig.p.min.hat, y=mu.p.min.hat, labels="GMIN Hat", col="orange", pos=2, cex=1.5) 
```

To sum up:
* For the global minimum variance portfolio we have roughly unbiased estimates of the weights, expected return and variance. 
* We have a fairly precise estimate of volatility but an imprecise estimate of expected return.

## True and estimated Sharpe ratios

For the example data, assume a monthly risk-free rate of $r_{f}=0.03/12=0.0025$. The true Sharpe ratios are:
```{r}
r.f = 0.03/12
SR.true = (mu.vec - r.f)/sd.vec 
SR.true
```

* Here, asset A has a much higher Sharpe ratio than asset B.

For the simulated data, the estimated Sharpe ratios are:

```{r}
SR.hat = (muhat.vals - r.f)/sigmahat.vals 
SR.hat
```

* For both assets the estimated Sharpe ratios are quite different from the true Sharpe ratios and are highly misleading.

## True and estimated Sharpe ratios

```{r}
plot(sd.vec, mu.vec, pch=16, col="black",       
     ylim=c(-0.01, 0.04), xlim=c(0, 0.1),      
     xlab=expression(sigma[p]), ylab=expression(mu[p]), cex=2) 
abline(h=0) 
points(0, r.f, pch=16, col="green", cex=2) 
text(0, r.f, labels=expression(r[f]), cex=1.5, pos=2) 
abline(a=r.f, b=SR.true[1], col="black") 
abline(a=r.f, b=SR.true[2], col="black")
text(x=sig.A, y=mu.A, labels="A", pos=4, cex = 1.5) 
text(x=sig.B, y=mu.B, labels="B", pos=4, cex = 1.5) 
# estimates 
points(sigmahat.vals, muhat.vals, pch=16, col="blue", cex=2) 
text(x=sigmahat.vals[1], y=muhat.vals[1], labels="A Hat",       
col="blue", pos=4, cex = 1.5) 
text(x=sigmahat.vals[2], y=muhat.vals[2], labels="B Hat",       
col="blue", pos=4, cex = 1.5) 
abline(a=r.f, b=SR.hat[1], col="blue") 
abline(a=r.f, b=SR.hat[2], col="blue")
```

* True Sharpe ratios are black lines
* Estimated Sharpe ratios are blue lines

## Bootstrap estimates of Sharpe Ratios

```{r}
n.boot = 500 
mu.boot = matrix(0, n.boot, 2) 
sd.boot = matrix(0, n.boot, 2) 
SR.boot = matrix(0, n.boot, 2) 
colnames(mu.boot) = colnames(sd.boot) = colnames(SR.boot) = c("A", "B")
set.seed(123) 
for (i in 1:n.boot) {   
   boot.idx = sample(n.obs, replace=TRUE)   
   ret.boot = returns.sim[boot.idx, ]    
   mu.boot[i, ] = colMeans(ret.boot)   
   sd.boot[i, ] = apply(ret.boot, 2, sd)    
   SR.boot[i, ] = (mu.boot[i, ] - r.f)/sd.boot[i, ] 
} 
# find index of max and min bootstrap Sharpe ratios
maxSRidx.A = which(SR.boot[, 1] == max(SR.boot[, 1])) 
minSRidx.A = which(SR.boot[, 1] == min(SR.boot[, 1])) 
maxSRidx.B = which(SR.boot[, 2] == max(SR.boot[, 2])) 
minSRidx.B = which(SR.boot[, 2] == min(SR.boot[, 2]))
```

## Bootstrap estimates of Sharpe Ratios

```{r, echo=FALSE}
# layout plot
plot(sd.vec, mu.vec, type="n",  ylim=c(-0.01, 0.06), xlim=c(0, 0.1),      
     xlab=expression(sigma[p]), ylab=expression(mu[p])) 
# bootstrap estimates 
points(sd.boot[, "A"], mu.boot[, "A"], pch=16, col="lightblue") 
points(sd.boot[, "B"], mu.boot[, "B"], pch=16, col="grey")
abline(h=0) 
# sample estimates 
points(sigmahat.vals, muhat.vals, pch=16, col="blue", cex=2) 
text(x=sigmahat.vals[1], y=muhat.vals[1], labels="A Hat",       
     col="blue", pos=4, cex = 1.5) 
text(x=sigmahat.vals[2], y=muhat.vals[2], labels="B Hat",       
     col="blue", pos=4, cex = 1.5)
# Min and max Sharpe ratios 
abline(a=r.f, b=SR.boot[maxSRidx.A, "A"], col="lightblue", lwd=2) 
abline(a=r.f, b=SR.boot[minSRidx.A, "A"], col="lightblue", lwd=2) 
abline(a=r.f, b=SR.boot[maxSRidx.B, "B"], col="grey", lwd=2) 
abline(a=r.f, b=SR.boot[minSRidx.B, "B"], col="grey", lwd=2) 
# Risk free rate 
points(0, r.f, pch=16, col="green", cex=2) 
text(0, r.f, labels=expression(r[f]), cex=1.5, pos=2)
```

* The maximum and minimum bootstrap Sharpe ratio estimates for asset B are the two grey straight lines from the risk-free rate that have the highest and lowest slopes, respectively. 
* The maximum and minimum bootstrap Sharpe ratio estimates for asset A are the two light blue lines with the highest and lowest slopes, respectively.

## Statistical properties of estimated Sharpe ratios

The bootstrap estimates of bias are:

```{r}
# bias
biasSR.boot = colMeans(SR.boot) - c(SR.true) 
biasSR.boot 
# bias relative to true values
biasSR.boot/SR.true
```

* Here we see there is a substantial upward bias in $\widehat{\mathrm{SR}}_{A}$ and a substantial downward bias in $\widehat{\mathrm{SR}}_{B}$

A crude bias adjusted Sharpe ratio estimate, which substracts the estimated bias from the sample estimate, gives results much closer to the true Sharpe ratios:

```{r}
SR.hat.biasAdjust = SR.hat - biasSR.boot 
SR.hat.biasAdjust
SR.true
```

The bootstrap estimated standard errors for both assets are big and indicate that the Sharpe ratios are not estimated well:

```{r}
apply(SR.boot, 2, sd)
```

## Bootstrap distribution of estimated Sharpe ratios

```{r, echo=FALSE}
par(mfrow=c(2,2))   
  hist(SR.boot[, 1], main="Asset A", xlab="bootstrap estimates of SR", col="slateblue1")   
  abline(v=SR.hat[1], col="white", lwd=2)   
  hist(SR.boot[, 2], main="Asset B", xlab="bootstrap estimates of SR", col="slateblue1")   
  abline(v=SR.hat[2], col="white", lwd=2)   
  qqnorm(SR.boot[, 1])   
  qqline(SR.boot[, 1])   
  qqnorm(SR.boot[, 2])   
  qqline(SR.boot[, 2]) 
par(mfrow=c(1,1))
```

 
## Three asset example data

Microsoft, Nordstrom and Starbucks monthly returns over the 5-year period January 1995 to December 1999.

```{r}
# get data and create monthly returns
data(msftDailyPrices, jwnDailyPrices, sbuxDailyPrices)
gwnDailyPrices = merge(msftDailyPrices, jwnDailyPrices, sbuxDailyPrices)
gwnMonthlyPrices = to.monthly(gwnDailyPrices, OHLC = FALSE)
gwnReturns = CalculateReturns(gwnMonthlyPrices)
gwnReturns = gwnReturns["1995::1999"]
colnames(gwnReturns)[2] = "NORD"
```

## Estimate GWN model parameters and calculate standard errors for means and volatilities

```{r}
# estimates
mu.hat = colMeans(gwnReturns)
sd.hat = apply(gwnReturns, 2, sd)
cov.hat = cov(gwnReturns)
cor.hat = cor(gwnReturns)
# estimated SE values for means and volatilities
n.obs = nrow(gwnReturns)
se.mu.hat = sd.hat/sqrt(n.obs)
se.sd.hat = sd.hat/sqrt(2*n.obs)
```

Estimates with standard errors

```{r}
# means with SE
rbind(mu.hat, se.mu.hat)
```

* Means are not estimated precisely

```{r}
# volatilities with SE
rbind(sd.hat, se.sd.hat)
```

* Volatilities are estimated more precisely


## Risk return characteristics without estimation error

```{r, echo=FALSE}
# show risk return tradeoffs
plot(sd.hat, mu.hat,  ylim=c(-0.03, 0.1), xlim=c(0, 0.20), ylab=expression(mu[p]),
     xlab=expression(sigma[p]), pch=16, col="blue", cex=2.5, cex.lab=1.75)     
text(sd.hat, mu.hat, labels=names(mu.hat), pos=4, cex = cex.val)
abline(h=0)

```

## Bootstrap means and volatilities to illustrate estimation error

```{r}
# re-sample means and sd values
n.boot = 500
mu.boot = matrix(0, n.boot, ncol(gwnReturns))
sd.boot = matrix(0, n.boot, ncol(gwnReturns))
colnames(mu.boot) = colnames(sd.boot) = colnames(gwnReturns)

set.seed(123)
for (i in 1:n.boot) {
  boot.idx = sample(n.obs, replace=TRUE)
  ret.boot = gwnReturns[boot.idx, ] 
  mu.boot[i, ] = colMeans(ret.boot)
  sd.boot[i, ] = apply(ret.boot, 2, sd) 
}

```

## Risk return characteristics with estimation error

```{r, echo=FALSE}
plot(sd.boot[, "SBUX"], mu.boot[, "SBUX"], col="grey", pch=16,
     ylim=c(-0.03, 0.10), xlim=c(0.0, 0.20),
     ylab=expression(mu[p]),
     xlab=expression(sigma[p]), cex.lab=1.5)
points(sd.hat["SBUX"], mu.hat["SBUX"], pch=16, col="black", cex=2.5)
abline(h=0)
text(sd.hat["SBUX"], mu.hat["SBUX"], labels="SBUX", pos=4, cex = 2)
# plot nordstrom
points(sd.boot[, "NORD"], mu.boot[, "NORD"], col="lightblue", pch=16)
points(sd.hat["NORD"], mu.hat["NORD"], pch=16, col="blue", cex=2.5)
text(sd.hat["NORD"], mu.hat["NORD"], labels="NORD", pos=4, cex = 2)
# plot microsoft
points(sd.boot[, "MSFT"], mu.boot[, "MSFT"], col="pink", pch=16)
points(sd.hat["MSFT"], mu.hat["MSFT"], pch=16, col="red", cex=2.5)
text(sd.hat["MSFT"], mu.hat["MSFT"], labels="MSFT", pos=4, cex = 2)
```

* Sample estimates are big dark circles
* Eootstrap estimates are small light circles
* Estimation error is more prounounced in means than volatilities

## Compute global minimum variance portfolio

```{r}
gmin.port = globalMin.portfolio(mu.hat, cov.hat)
gmin.port
```

## Risk return characteristics without estimation error

```{r, echo=FALSE}
plot(sd.hat, mu.hat,  ylim=c(-0.03, 0.1), xlim=c(0, 0.20), ylab=expression(mu[p]),
     xlab=expression(sigma[p]), pch=16, col="blue", cex=2.5, cex.lab=1.75)     
text(sd.hat, mu.hat, labels=names(mu.hat), pos=4, cex = cex.val)
points(gmin.port$sd, gmin.port$er, pch=16, cex=2.5, col="green")
text(gmin.port$sd, gmin.port$er, labels="Global Min", pos=2, cex = cex.val)
abline(h=0)
```

## Bootstrap global minimum variance portfolio

```{r}
# bootstrap global min portfolio
mu.gmin.boot = matrix(0, n.boot, 1)
sd.gmin.boot = matrix(0, n.boot, 1)
w.gmin.boot = matrix(0, n.boot, 3)
colnames(mu.gmin.boot) = colnames(sd.gmin.boot) = "global.min"
colnames(w.gmin.boot) = names(mu.hat)

set.seed(123)
for (i in 1:n.boot) {
  boot.idx = sample(n.obs, replace=TRUE)
  ret.boot = gwnReturns[boot.idx, ] 
  mu.boot = colMeans(ret.boot)
  cov.boot = cov(ret.boot) 
  gmin.boot = globalMin.portfolio(mu.boot, cov.boot)
  mu.gmin.boot[i, ] = gmin.boot$er
  sd.gmin.boot[i, ] = gmin.boot$sd
  w.gmin.boot[i, ] = gmin.boot$weights
}

```

## Risk return characteristics without estimation error

```{r, echo=FALSE}
plot(sd.hat, mu.hat,  ylim=c(-0.03, 0.1), xlim=c(0, 0.20), ylab=expression(mu[p]),
     xlab=expression(sigma[p]), pch=16, col="blue", cex=2.5, cex.lab=1.75)     
abline(h=0, v=0)
text(sd.hat, mu.hat, labels=names(mu.hat), pos=4, cex = cex.val)
# plot bootstrapped global min
points(sd.gmin.boot, mu.gmin.boot, col="lightgreen", pch=16)
points(gmin.port$sd, gmin.port$er, pch=16, cex=2.5, col="black")
text(gmin.port$sd, gmin.port$er, labels="Global Min", pos=2, cex = cex.val)
```

* Estimation error is larger in mean than in volatility (cloud is more stretched in vertical direction)

## Bootstrap distribution of global minimum variance portfolio mean and volatility 

```{r, echo=FALSE}
# look at bootstrap distribution
par(mfrow=c(2,2))
  hist(mu.gmin.boot, col="cornflowerblue")
  qqnorm(mu.gmin.boot, col="cornflowerblue", pch=16)
  qqline(mu.gmin.boot)

  hist(sd.gmin.boot, col="cornflowerblue")
  qqnorm(sd.gmin.boot, col="cornflowerblue", pch=16)
  qqline(sd.gmin.boot)
par(mfrow=c(1,1))
```

* Both distribution look pretty normal
* histogram of mean is more spread out (0% - 7%) than histogram of volatility (5% - 9%)

## Bootstrap bias, SE and 95% CI for mean

```{r}
bias.mu.gmin = mean(mu.gmin.boot) - gmin.port$er
se.mu.gmin = sd(mu.gmin.boot)
ci.mu.gmin.95 = c(gmin.port$er-2*se.mu.gmin, 
                  gmin.port$er+2*se.mu.gmin)
names(ci.mu.gmin.95) = c("lower", "upper")
```

Bootstrap estimate of bias for mean is small

```{r}
bias.mu.gmin
```

Bootstrap SE for mean is fairly big

```{r}
se.mu.gmin
```

Bootstrap 95% CI for mean is pretty wide

```{r}
ci.mu.gmin.95
```

## Bootstrap bias, SE and 95% CI for volatility

```{r}
bias.sd.gmin = mean(sd.gmin.boot) - gmin.port$sd
se.sd.gmin = sd(sd.gmin.boot)
ci.sd.gmin.95 = c(gmin.port$sd-2*se.sd.gmin, 
                  gmin.port$sd+2*se.sd.gmin)
names(ci.sd.gmin.95) = c("lower", "upper")
```

Bootstrap estimate of bias for volatility is small

```{r}
bias.sd.gmin
```

Bootstrap SE for volatility is fairly small

```{r}
se.sd.gmin
```

Bootstrap 95% CI for volatility is not so wide

```{r}
ci.sd.gmin.95
```

## Bootstrap weights in global minimum variance portfolio

```{r, echo=FALSE}
par(mfrow=c(2,2))
  hist(w.gmin.boot[, "SBUX"], main="SBUX", xlab="Weight", col="cornflowerblue")
  hist(w.gmin.boot[, "MSFT"], main="MSFT", xlab="Weight", col="cornflowerblue")
  hist(w.gmin.boot[, "NORD"], main="NORD", xlab="Weight", col="cornflowerblue")
  corrplot(cor(w.gmin.boot), method="ellipse")
par(mfrow=c(1,1))
```

* histograms of weights are faily spread out
* Weights are negatively correlated (because they sum to one)

## Bootstrap bias, SE and 95% CI for weights

```{r}
bias.w.gmin = colMeans(w.gmin.boot) - gmin.port$weights
se.w.gmin = apply(w.gmin.boot, 2, sd)
ci.w.gmin.95 = rbind(gmin.port$weights-2*se.w.gmin, 
                     gmin.port$weights+2*se.w.gmin)
rownames(ci.w.gmin.95) = c("lower", "upper")
```

Bootstrap estimates of bias for weights are fairly small

```{r}
bias.w.gmin
```

Bootstrap SEs for weights are moderately large (9%)

```{r}
se.w.gmin
```

Bootstrap 95% CIs for weights are wide

```{r}
ci.w.gmin.95
```


## Bootstrap weights in global minimum variance portfolio

```{r, echo=FALSE}
# sort bootstrap values by mean
tmp.w.boot = w.gmin.boot[1:20, ]
tmp.mu.boot = mu.gmin.boot[1:20, ]
tmp.sd.boot = sd.gmin.boot[1:20, ]
sort.idx = order(tmp.mu.boot)

# look at weights in stacked bar charts
chart.StackedBar(tmp.w.boot[sort.idx,], 
                 xaxis.labels=round(tmp.mu.boot[sort.idx],digits=3), 
                 xlab="Portfolio SD", ylab="Weights", cex.lab=1.5,
                 cex.axis=1.5)
```

* x-axis is bootstrap mean value of global minimum variance portfolio
* weights vary a bit from portfolio to portfolio - SBUX always has the smallest weight

## Efficient frontier of risky assets

```{r}
# compute efficient frontier
ef = efficient.frontier(mu.hat, cov.hat)
```

```{r, echo=FALSE}
# plot efficient portfolios
plot(ef$sd, ef$er, type="b", ylim=c(-0.03, 0.1), xlim=c(0, 0.20), 
     pch=16, col="blue", cex=2, ylab=expression(mu[p]), xlab=expression(sigma[p]))

points(sd.hat, mu.hat, pch=16, cex=2, col="black")
points(gmin.port$sd, gmin.port$er, pch=16, cex=2, col="green")
text(sd.hat, mu.hat, labels=names(mu.hat), pos=4, cex=2)
text(gmin.port$sd, gmin.port$er, labels="Global min", pos=4, cex=2)
```

## Bootstrap efficient frontier

```{r}
ef.list = list()
set.seed(123)
for (i in 1:n.boot) {
  boot.idx = sample(n.obs, replace=TRUE)
  ret.boot = gwnReturns[boot.idx, ] 
  mu.boot = colMeans(ret.boot)
  cov.boot = cov(ret.boot) 
  ef.boot = efficient.frontier(mu.boot, cov.boot)
  ef.list[[i]] = ef.boot
}

```

## Show bootstrap frontier portfolios

```{r, echo=FALSE}
plot(ef$sd, ef$er, type="n", ylim=c(-0.03, 0.1), xlim=c(0, 0.20), 
     ylab=expression(mu[p]), xlab=expression(sigma[p]))
# show bootstrap frontiers
for (i in 1:n.boot) {
  points(ef.list[[i]]$sd, ef.list[[i]]$er, type="b",
         pch=16, col="grey", cex=1.5)
}
# show sample frontier and assets
points(ef$sd, ef$er, type="b", pch=16, col="blue", cex=2)
points(sd.hat, mu.hat, pch=16, cex=2, col="black")
text(sd.hat, mu.hat, labels=names(mu.hat), pos=4, cex=2)

```

## A prettier picture with 20 portfolios

```{r, echo=FALSE}
plot(ef$sd, ef$er, type="n", ylim=c(-0.03, 0.1), xlim=c(0, 0.20), 
     ylab=expression(mu[p]), xlab=expression(sigma[p]))
# show bootstrap frontiers
for (i in 1:20) {
  points(ef.list[[i]]$sd, ef.list[[i]]$er, type="b",
         pch=16, col=i, cex=1)
}
# show sample frontier and assets
points(ef$sd, ef$er, type="b", pch=16, col="blue", cex=2)
points(sd.hat, mu.hat, pch=16, cex=2, col="black")
text(sd.hat, mu.hat, labels=names(mu.hat), pos=4, cex=2)

```
